A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-08-25 main.tex
4.4 and 5 edits

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n << p$, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. When $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$. LMM-Lasso-ggmix achieves a moderate improvement in the estimation of $\eta$, particularly in the presence of lower $\eta$ values and where $n << p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. Interestingly, these improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix. When $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ in suffers. When $n \ge 1000$ and in the presence of substantial environmental confounding, $\xi = 0.8$,  LMM-Lasso-ggmix's is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso Rakitsch outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ where $n \ge p$, likely due to the ill-conditioning of the RRM in these settings \anna{cite French paper (?)}. 
% Given the intended use of these methods for the analysis of genetic data, where the number of genetic variants often outnumber the number of subjects, these results could be of little consequence to practitioners. 
When $n < p$, LMM-Lasso-ggmix estimates $\bbeta$ as well as LMM-Lasso-Rakitsch, and $\eta$ better than LMM-Lasso-Rakitsch. However, in this setting the computational time LMM-Lasso-ggmix reqiures may be prohibitive for practical applications. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required for each method in the various simulation settings.

\anna{ below in progress}

Unfortunately, the settings where ggmix estimates things well coincide with excessive times and when it works well in terms of time, it estimates poorly

Despite its comparable performance in accurately estimating $\bbetaHat$, LMM-Lasso-ggmix takes requires an average of \anna{73.4} times the computing time of LMM-Lasso-Rakitsch. Thus, LMM-Lasso-Rakitsch seems to be a more attractive and scalable method than LMM-Lasso-ggmix.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion} \label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\anna{this to change somewhat based on time/convergence results...}

@@

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from instability in low dimensions. \anna{depends on ggmix time results} Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably in high dimensions, LMM-Lasso-ggmix requires an average of 73.4 and as much as 167.57 \anna{73.4 is the mean ratio + sd, but the max absolute ratio is 167.57 - which should I report?} times the computing time of LMM-Lasso-Rakitsch due to its iterative estimation procedure. Thirdly, while the additional time required of LMM-Lasso-ggmix may be worthwhile if it yielded more accurate estimates of $\eta$, or narrow-sense heritability, we have been unable to identify a scenario where its narrow-sense heritability estimates are greatly superior to those of LMM-Lasso-Rakitsch. We have implemented the LMM-Lasso-Rakitsch method using the equivalent, but more interpretable, $\eta$ paramaterization in the \texttt{R} package, \texttt{penalizedLMM (final name tbd)}, available on the author's github page: \url{https://github.com/areisett/penalizedLMM}.

\pb{Introduce penalizedLMM here, or in next paper?}

>>>>

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n << p$, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. However, when $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$.

LMM-Lasso-ggmix achieves a moderate improvement in the estimation of $\eta$ compared to LMM-Lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n << p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. These improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix. In addition, when $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ suffers. When $n \ge 1000$ and in the presence of substantial environmental confounding, $\xi = 0.8$,  LMM-Lasso-ggmix's performance is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n << p$. However, these are also the settings in which LMM-Lasso-ggmix is most computationally inefficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n << p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch.

Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently, likely due to the ill-conditioning of the RRM \anna{cite French paper (?)}.

% Given the intended use of these methods for the analysis of genetic data, where the number of genetic variants often outnumber the number of subjects, these results could be of little consequence to practitioners.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion} \label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus considered by the authors to be more broadly applicable.

\citet{Rakitsch2012} parameterize their model in terms of the variance components $\sigmagg$ and $\delta := \sigmaee / \sigmagg$ as previously discussed, however, we have implemented the LMM-Lasso-Rakitsch method using an equivalent but more interpretable parameterization in terms of $\eta$ and $\sigmaee$ in the \texttt{R} package, \texttt{penalizedLMM}, available on the author's github repository \url{https://github.com/areisett/penalizedLMM}.
